Likelihoods for a general class of ARGs under the SMC

### 2025-04-28 review-responses.tex

%%%%%%%%%%%%%%
\reviewersection{2}
\begin{quote}
This paper does a couple of interrelated things.
\begin{enumerate}
    \item The authors develop a new backward-in-time formulation of the Sequentially Markov Coalescent (SMC) model.
    \item This backward-in-time formulation of the SMC model provides a way to efficiently calculate likelihood for a sample of genomes from this model, provided that we have information on how these genomes are related to each other. The latter is done by a data structure called Ancestral Recombination Graph (ARG). The authors provide an algorithm and Python code for this likelihood calculation (is this emphasised enough across the paper!?).
    \item The authors also evaluate the likelihood under different levels of information available in the ARG. This enables them to connect challenges when comparing methods and outputs from the rigorous, yet ``scale-limited'', probabilistic ARG inference methods with scaleable, yet less-rigorous, heuristic ARG inference methods.
\end{enumerate}
I welcome this paper and quite like it's succinctness. I am not able to follow all the maths, but the rationale and presentation is clear to me and I only have minor comments.
\end{quote}

% ABSTRACT

\begin{point}{\revref} % \&75
 Isn't it possible to run SINGER on up to thousands of samples? I know that several hundreds is also a thousand!
\end{point}

\reply{
We think that the characterisation of SINGER being feasible for hundreds of
samples is fair and reasonable. In the SINGER preprint, the maximum sample 
size evaluated on simulations is 300, and 
the flagship real data example 
uses only 200 African samples from the 1000 Genomes dataset.
How we define ``feasible''  is of course subjective, but 
the scale of inferences done in the 
method's paper seems like a reasonable approximation.
}

### 2025-04-28 paper.tex
Hit a few comments

In Fig~\ref{fig:lik-surface} we begin with a realisation of the SMC
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik}
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$.
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we
simplify out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes
contain all the information that we need about recombination events
for this application.
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  \revpoint{2}{31}
we can see there is a significant loss of information, \revpoint{2}{32}
and the minimum of the
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying
degrees of ARG simplification.)

Fig~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig~\ref{fig:lik-surface},
and even on the fully simplified
ARG has only a very minor effect on location of the minimum.

Fig~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the
\cite{li2003modeling} copying process. Information about
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to
attempt to reconstruct the exact subtree-prune-and-regraft
moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well \revpoint{2}{37}
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we
fully simplify the output of tsinfer
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a significant bias \revpoint{2}{38}
away from the true value of the recombination rate parameter.

>>>>

In Fig.~\ref{fig:lik-surface} we begin with a realisation of the SMC
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik}
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$.
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we
simplify out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes
contain all the information that we need about recombination events
for this application.
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  \revpoint{2}{31}
we can see there is a significant loss of information, \revpoint{2}{32}
and the minimum of the
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying
degrees of ARG simplification.)

Fig.~\ref{fig:lik-surface} shows that Eq.~\eqref{eq:full-lik} is also remarkably \revpoint{2}{33}
robust to the presences of polytomies in this example. For each of the three
ARGs discussed in the previous paragraph, we also evaluated the likelihood
for ARGs in which 5\% or 25\% of the internal ARG nodes are removed and
edges adjusted accordingly.
This introduces polytomies at varying degrees in the different ARGs because
deleting a node in a tree (and connecting its children directly to its
parent) will only create a polytomy if the parent
is not unary. Thus, while deleting 25\% of the internal nodes
in the fully simplified ARG created an average of 33 polytomies
over the 2,933 trees along the sequence (each tree having
200 leaf nodes), it resulted in an average of 21 polytomies
per tree in the ARG where recombination nodes have been removed,
and only 10 in the full ARG (which has an average of 1026 unary
nodes per tree).
%     name                  leaves  unary       binary  poly
% 0   full_arg_smc          200.0   1094.417661 199.000000  0.000000
% 1   unary_coal            200.0   456.396522  199.000000  0.000000
% 2   fully_simplified      200.0   0.000000    199.000000  0.000000
% 3   tsinfer_tsdate        200.0   192.447590  119.754399  32.270849
% 4   tsinfer_fs_tsdate     200.0   0.000000    119.563898  32.322684
% 5   py_full_arg_smc_0.05  200.0   1025.565973 192.662462  3.153086
% 6   py_unary_coal_0.05    200.0   429.315017  190.182253  4.372696
% 7   py_fully_simied_0.05  200.0   0.000000    185.691468  6.393174
% 8   py_full_arg_smc_0.25  200.0   779.846233  176.830549  10.307876
% 9   py_unary_coal_0.25    200.0   323.054926  154.422588  21.104016
% 10  py_fully_simied_0.25  200.0   0.000000    117.096174  33.241986
Given this disparity in the number of polytomies introduced, the
effects on the likelihood across the three examples are not entirely comparable.
For the ARGs containing unary nodes,
removing 25\% of the internal
nodes has negligible effect on the likelihood curve in Fig.~\ref{fig:lik-surface},
and even on the fully simplified
ARG has only a very minor effect on location of the minimum.

Fig.~\ref{fig:lik-surface} also shows results for ARGs inferred from the  \revpoint{2}{35}
simulated data using tsinfer and tsdate~\citep{wohns_unified_2022},
using default parameters for tsinfer, and the true values of mutation
rate and $N_e$ for tsdate.
In the first, we use the default output of tsinfer
(2,148 nodes; 5,589 edges; 1,307 trees; 3,039 mutations). \revpoint{2}{36}
This ARG has an average of 192 unary nodes, 120 binary nodes, and 32 nodes
with three or more children per tree. Tsinfer does not directly estimate
recombination events; the breakpoints between trees are the consequence of
switches between parental haplotypes in the
\cite{li2003modeling} copying process. Information about
recombination events in the returned ARG is therefore quite imprecise,
and likely to contain incorrect and contradictory details if one were to
attempt to reconstruct the exact subtree-prune-and-regraft
moves~\citep[e.g.,][]{rasmussen2023espalier}. Nonetheless, in
this toy example at least, our likelihood function performs well \revpoint{2}{37}
on the output of tsinfer, with the minimum close to the true
parameter value. We also show the likelihood curve obtained when we
fully simplify the output of tsinfer
(1,865 nodes; 7,511 edges; 1,252 trees; 3,039 mutations)
before dating. We can see that, as with the true data,
removing all the unary nodes results in a bias \revpoint{2}{38}
away from the true value of the recombination rate parameter.
